
\section{Course progress} 

\begin{multicols}{3}
    \begin{itemize}
        \item First reading
        \begin{todolist}
            % \item[\done] Frame the problem
            \item[\done] Lecture 1-22
            \item Lecture 23
            \item Lecture 24
            \item Lecture 25
        \end{todolist}
        \item Gjennomgang
        \begin{todolist}
            \item Lecture[\done] 1-2
            \item Lecture 3-4
            \item Lecture 5-6
            \item Lecture 7-8
            \item Lecture 9-10
            \item Lecture 11-12
            \item Lecture 13-14
            \item Lecture 15-16
            \item Lecture 17-18
            \item Lecture 19-20
            \item Lecture 21-22
            \item Lecture 23
        \end{todolist}
    \end{itemize}
\end{multicols}


\subsection{Exams}
\begin{multicols}{3}
    \begin{todolist}
        \item May 2023
        \item June 2019
        \item May 2018
        \item May 2017
        \item June 2016
        \item May 2015
        \item May 2014
        \item August 2014    
    \end{todolist}        
\end{multicols}

\subsection{Topics}
\subsubsection{1}
Multivariate distributions and expectations (HS 4.1-4.2).

Multivariate moments (HS 4.2 using HS 2.1-2.4).

Transformations (HS 4.3, 4.4)

PCA (HS 11.1-11.3).

Charactestic functions (HS 4.2). 


\subsubsection{2}
Multivariate normal distribution (HS 4.4, 5.1).

Estimation in the multivariate normal distribution (HS 3.3, 4.5).

Quadratic forms and idempotent matrices (FKLM Appendix B, Th. B2, B8).


\subsubsection{3}
Multiple linear regression: model, parameter estimation (FKLM 3.1, 3.2).

Properties of estimators, fitted values, residuals (FKLM 3.2). 

Inference about coefficients (FKLM 3.3).

Multiple linear regression: t-test about coefficients, ANOVA decomposition, coefficient of determination, F-test (FKLM 3.2, 3.3).

General F-test for regression coefficients (FKLM 3.2, 3.3, 3.5).

transformation of data (FKLM 3.2, 3.3, 3.4, 3.5).



\subsubsection{4}
Model analysis and model selection (FKLM 3.4).

Multiple hypothesis testing (HBL).

Examples.


\subsubsection{5}
ANOVA (HS 8.1.1). 

Design of experiment (DOE): two-level factorial design (T).

\subsection{Keywords to know}
